---
title: "Bee Diversity Analysis using MeanRarity"
format: html
editor: visual
---

# Analyzing Bee Diversity in a BACI Study Design

This analysis uses the methods described in Roswell et al. (2021) to analyze bee diversity across multiple sites in a Before-After-Control-Impact (BACI) design. We'll use coverage standardization and Hill diversity metrics to provide a robust comparison of bee diversity.

## Setup

First, we'll load the necessary packages and read in our data.

```{r setup}
#| message: false
#| warning: false



# Load required packages

library(MeanRarity)  # For calculating Hill diversity metrics
library(iNEXT)       # For coverage-based rarefaction
library(dplyr)       # For data manipulation
library(tidyr)       # For reshaping data
library(ggplot2)     # For visualization
library(lubridate)   # For date handling
library(lme4)        # For mixed-effects models
library(lmerTest)    # For p-values in mixed models
library(emmeans)     # For estimated marginal means

# Read in the data
bee_data <- read.csv("../data/processed/bee_02-18_data.csv") 
bee_traits <- read.csv("../data/processed/bee_traits.csv", header = TRUE)
site_data <- read.csv("../data/processed/sites.csv", header = TRUE)
dates_to_drop <- read.csv("../data/processed/02_03_18_datestodrop.csv", header = TRUE) 
```

## Data Cleaning

Next, we'll clean up the data, standardizing site names and formatting dates.

```{r data_cleaning}
#| message: false
#| warning: false
#| echo: false

# Fix site names
bee_data_clean <- bee_data %>%
  mutate(site = case_when(
    site %in% c("Wappo Hill", "Wappo hill") ~ "Wappo", 
    site %in% c("Stag's leap", "Stags Leap", "Stag's Leap") ~ "Stags", 
    site %in% c("Veteran's Home") ~ "Veterans", 
    TRUE ~ site
  ))

# Reformat dates
bee_data_clean <- bee_data_clean %>% 
  mutate(start_date = mdy(start_date))
dates_to_drop <- dates_to_drop %>% 
  mutate(date = mdy(date))

# Create combined species name
bee_data_clean$combined_name <- paste(bee_data_clean$genus_name, bee_data_clean$species, sep=" ")

# Look at data structure
str(bee_data_clean)
```

## Preparing Data for Diversity Analysis

To calculate Hill diversity metrics, we need to create a matrix of species abundances per site and year.

```{r create_abundance_matrix}
#l warning = FALSE
#l echo = FALSE
# Create a wide-format abundance matrix for diversity analysis
# Each row is a site-year combination, columns are species
abundance_matrix <- bee_data_clean %>%
  group_by(site, year, combined_name) %>%
  # Sum abundance of each species at each site-year
  summarize(abundance = sum(number), .groups = "drop") %>%
  # Convert to wide format: species as columns
  pivot_wider(
    id_cols = c(site, year),
    names_from = combined_name,
    values_from = abundance,
    values_fill = 0  # Fill missing species with 0
  )

# Safely extract site_year_ids with base R
site_year_ids <- data.frame(
  site = abundance_matrix$site,
  year = abundance_matrix$year
)

# Safely create abundance_values matrix
# This avoids issues with special characters in species names
abundance_values <- as.matrix(abundance_matrix[, 3:ncol(abundance_matrix)])

# Check dimensions
cat("Number of site-year combinations:", nrow(site_year_ids), "\n")
cat("Number of species columns:", ncol(abundance_values), "\n")

# Add metadata about sampling effort
sampling_effort <- bee_data_clean %>%
  group_by(site, year) %>%
  summarise(sampling_events = n_distinct(start_date))

# Add treatment information from site_data
site_metadata <- site_year_ids %>%
  left_join(sampling_effort, by = c("site", "year")) %>%
  left_join(site_data %>% dplyr::select(site, treatment), by = "site")

head(site_metadata)
```

## Coverage-based Standardization

According to Roswell et al. (2021), coverage is a more robust way to standardize samples than equal effort or equal sample size. Coverage estimates what proportion of the total community (including undetected species) is represented in our sample.

```{r calculate_coverage}
# Calculate coverage for each site-year combination
site_coverage <- data.frame(site_year_ids, 
                           coverage = apply(abundance_values, 1, function(x) {
                             # Count singletons (species with only 1 individual)
                             f1 <- sum(x == 1)
                             # Get total number of individuals
                             n <- sum(x)
                             # Calculate coverage using the formula from Chao and Jost (2012)
                             if (f1 == 0) {
                               return(1) # Perfect coverage if no singletons
                             } else {
                               return(1 - (f1/n) * ((n-1)/n))
                             }
                           }))

# Find the minimum coverage across all site-years (to use as standardization point)
min_coverage <- min(site_coverage$coverage)
cat("Minimum coverage across all site-years:", min_coverage, "\n")

# Display coverage by site and year
site_coverage %>%
  arrange(coverage) %>%
  knitr::kable(caption = "Sample coverage by site and year")
```

## Calculating Hill Diversity Metrics

Hill diversity unifies different diversity metrics (richness, Shannon, Simpson) into a common framework. The parameter ℓ (or q in some literature) determines how much weight is given to rare versus common species.

The MeanRarity package uses parameter q in its code implementation, while their conceptual framework in the paper uses ℓ. The relationship between them is ℓ = 1-q, so:

Species richness: ℓ=1 corresponds to q=0 Hill-Shannon: ℓ=0 corresponds to q=1 Hill-Simpson: ℓ=-1 corresponds to q=2

```{r calculate_diversity}
#| warning: false
# Function to calculate Hill diversity metrics for each site-year
calculate_diversity_metrics <- function() {
  diversity_results <- data.frame()
  
  for (i in 1:nrow(site_year_ids)) {
    site_id <- site_year_ids$site[i]
    yr <- site_year_ids$year[i]
    
    # Extract abundance data for this site-year
    site_data_vec <- abundance_values[i, ]
    site_data_vec <- site_data_vec[site_data_vec > 0]  # Remove zeros
    
    # Calculate raw Hill diversity using the rarity function from MeanRarity
    richness_raw <- rarity(site_data_vec, q = 0)
    hill_shannon_raw <- rarity(site_data_vec, q = 1)
    hill_simpson_raw <- rarity(site_data_vec, q = 2)
    
    # Use iNEXT to estimate diversity at standardized coverage
    inext_result <- estimateD(site_data_vec, datatype = "abundance", 
                             base = "coverage", level = min_coverage)
    
    # The issue is in extracting values from inext_result
    # Print the structure to understand how to access the values
    # print(str(inext_result))
    
    # Get the values in the correct way
    # iNEXT results are sorted by order (0, 1, 2)
    richness_std <- inext_result$qD[1]     # First row is q=0 (richness)
    hill_shannon_std <- inext_result$qD[2] # Second row is q=1 (Shannon)
    hill_simpson_std <- inext_result$qD[3] # Third row is q=2 (Simpson)
    
    # Combine with existing metadata
    result_row <- data.frame(
      site = site_id,
      year = yr,
      raw_richness = richness_raw,
      raw_hill_shannon = hill_shannon_raw,
      raw_hill_simpson = hill_simpson_raw,
      std_richness = richness_std,
      std_hill_shannon = hill_shannon_std,
      std_hill_simpson = hill_simpson_std,
      coverage = site_coverage$coverage[i]
    )
    
    # Append to the result data frame
    diversity_results <- rbind(diversity_results, result_row)
  }
  
  # Add metadata about treatment and sampling effort
  diversity_results <- diversity_results %>%
    left_join(site_metadata, by = c("site", "year"))
  
  return(diversity_results)
}

# Calculate diversity metrics
diversity_results <- calculate_diversity_metrics()

# Create factors for time period and ensure proper ordering
diversity_results$period <- ifelse(diversity_results$year == "2018", "After", "Before")
diversity_results$period <- factor(diversity_results$period, levels = c("Before", "After"))
diversity_results$treatment <- factor(diversity_results$treatment, levels = c("Control", "Impact"))
diversity_results$year <- factor(diversity_results$year, levels = c("2002", "2003", "2018"))
```

## Visualizing Diversity Metrics

Now let's visualize the diversity results to better understand the patterns across sites, years, and treatments.

```{r visualize_diversity}
#| fig.width: 10
#| fig.height: 6

# Visualize Hill-Shannon diversity (standardized by coverage)
ggplot(diversity_results, aes(x = year, y = std_hill_shannon, color = treatment, group = site)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  labs(title = "Hill-Shannon Diversity (Coverage-Standardized)",
       x = "Year", 
       y = "Hill-Shannon Diversity (q=1)") +
  theme_minimal()

# Compare all three diversity metrics
diversity_long <- diversity_results %>%
  pivot_longer(cols = c(std_richness, std_hill_shannon, std_hill_simpson),
               names_to = "metric", 
               values_to = "value") %>%
  mutate(metric = factor(metric, 
                       levels = c("std_richness", "std_hill_shannon", "std_hill_simpson"),
                       labels = c("Species Richness (q=0)", 
                                 "Hill-Shannon (q=1)", 
                                 "Hill-Simpson (q=2)")))

ggplot(diversity_long, aes(x = year, y = value, color = treatment, group = site)) +
  facet_wrap(~ metric, scales = "free_y") +
  geom_point(size = 2) +
  geom_line(linewidth = 1) +
  labs(title = "Coverage-Standardized Hill Diversity Metrics",
       x = "Year", 
       y = "Diversity Value") +
  theme_minimal()
```

## Diversity Profiles

Diversity profiles show how diversity changes with the Hill exponent ℓ, giving a more complete picture of community structure.

```{r diversity_profiles}
#| fig.width: 8
#| fig.height: 6
#| warning: false

# Create diversity profiles for each site in the final year (2018)
# Filter to just 2018 data
data_2018 <- abundance_matrix %>% 
  filter(year == "2018")
site_ids_2018 <- data_2018$site

# Safely extract abundance data for 2018 sites
# Using the same approach we used above
abundance_2018 <- as.matrix(data_2018[, 3:ncol(data_2018)])

# Get treatment information
treatments_2018 <- site_metadata %>% 
  filter(year == "2018") %>% 
  dplyr::select(site, treatment)

# Set up the plot
par(mfrow = c(1, 1))
plot(NULL, xlim = c(-1, 1), ylim = c(0, max(diversity_results$std_richness) * 1.1),
     xlab = "Hill exponent (ℓ)", ylab = "Hill Diversity",
     main = "Diversity Profiles (2018)")

# Define colors for treatments
colors <- c("Control" = "blue", "Impact" = "red")

# Calculate and plot diversity profiles
for (i in 1:nrow(abundance_2018)) {
  site_id <- site_ids_2018[i]
  site_abund <- abundance_2018[i, ]
  site_abund <- site_abund[site_abund > 0]  # Remove zeros
  
  # Get treatment for this site
  site_treatment <- treatments_2018$treatment[treatments_2018$site == site_id]
  
  # Calculate diversity for a range of exponents
  exponents <- seq(-1, 1, by = 0.1)
  profile_values <- sapply(exponents, function(ell) rarity(site_abund, q = 1 - ell))
  
  # Plot the profile
  lines(exponents, profile_values, col = colors[site_treatment], lwd = 2)
  points(c(-1, 0, 1), profile_values[c(1, 11, 21)], 
        col = colors[site_treatment], pch = 19)
  text(1, profile_values[21], site_id, pos = 4, cex = 0.8)
}

# Add a legend
legend("topright", legend = names(colors), col = colors, lwd = 2, pch = 19)
```

## Statistical Analysis

Now we analyze the diversity data using linear mixed-effects models with a BACI design. As recommended by Roswell et al., we focus primarily on Hill-Shannon diversity for our statistical analysis because it provides a balanced measure of diversity.

```{r statistical_analysis}
# Model for Hill-Shannon diversity (recommended for ecological applications)
model_hill_shannon <- lmer(std_hill_shannon ~ year * treatment + (1 | site), 
                          data = diversity_results)
summary(model_hill_shannon)

# Extract p-values
p_values <- summary(model_hill_shannon)$coefficients[, "Pr(>|t|)"]
print(p_values)

# Calculate estimated marginal means
emm_hill_shannon <- emmeans(model_hill_shannon, ~ year * treatment)
pairs(emm_hill_shannon)

# Also analyze richness and Simpson diversity for comparison
model_richness <- lmer(std_richness ~ year * treatment + (1 | site), 
                      data = diversity_results)
summary(model_richness)

model_simpson <- lmer(std_hill_simpson ~ year * treatment + (1 | site), 
                     data = diversity_results)
summary(model_simpson)
```

```{r plot_emm, fig.width=8, fig.height=6}
# Plot the estimated means with error bars
emm_data <- as.data.frame(emm_hill_shannon)
ggplot(emm_data, aes(x = year, y = emmean, color = treatment, group = treatment)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = emmean - SE, ymax = emmean + SE), width = 0.2) +
  labs(title = "Hill-Shannon Diversity by Treatment and Year",
       x = "Year", 
       y = "Estimated Hill-Shannon Diversity") +
  theme_minimal()
```

## Analysis of Raw vs. Coverage-standardized Results

Let's compare the raw diversity metrics with the coverage-standardized metrics to see how standardization affects our conclusions.

The shape of the curves are similar although estimated species of the Hill Shannon estimate is lower.

```{r raw_vs_std, fig.width=10, fig.height=6}
# Combine raw and standardized metrics for comparison
diversity_compare <- diversity_results %>%
  dplyr::select(site, year, treatment, period, 
         raw_hill_shannon, std_hill_shannon,
         sampling_events, coverage) %>%
  pivot_longer(
    cols = c(raw_hill_shannon, std_hill_shannon),
    names_to = "metric_type",
    values_to = "hill_shannon"
  ) %>%
  mutate(metric_type = factor(metric_type, 
                            levels = c("raw_hill_shannon", "std_hill_shannon"),
                            labels = c("Raw Hill-Shannon", "Coverage-standardized Hill-Shannon")))

# Plot the comparison
ggplot(diversity_compare, aes(x = year, y = hill_shannon, linetype = treatment, color = site, group = site)) +
  facet_wrap(~ metric_type) +
  geom_point(size = 3) +
  geom_line(size = 1) +
    labs(title = "Raw vs. Coverage-standardized Hill-Shannon Diversity",
       x = "Year", 
       y = "Hill-Shannon Diversity") +
  theme_minimal()

# Check correlation between sampling events and raw diversity
cor_sampling_raw <- cor(diversity_results$sampling_events, diversity_results$raw_hill_shannon)
cat("Correlation between sampling events and raw Hill-Shannon diversity:", cor_sampling_raw, "\n")

# Check correlation between sampling events and standardized diversity
cor_sampling_std <- cor(diversity_results$sampling_events, diversity_results$std_hill_shannon)
cat("Correlation between sampling events and standardized Hill-Shannon diversity:", cor_sampling_std, "\n")
```

### Key Findings

1.  **Coverage Standardization**: Standardizing by coverage rather than by sample size or effort allows for more robust comparisons across sites and years. The minimum coverage across our samples was 0.769, which means we're comparing equally complete samples from each community.

2.  **Hill Diversity Metrics**: The three Hill diversity metrics (richness, Shannon, and Simpson) provide different perspectives on community diversity:

    -   Species richness (ℓ=1, q=0) gives equal weight to all species and is most sensitive to rare species
    -   Hill-Shannon diversity (ℓ=0, q=1) provides a balanced view of diversity
    -   Hill-Simpson diversity (ℓ=-1, q=2) emphasizes common species

3.  **BACI Analysis**: Our statistical analysis using a BACI design reveals:

    -   None of the models are significant
    -   We have no interaction terms that are significant
    -   Using the Simpson diversity which emphasizes common diversity, there is a difference between 2002 and 2018

### Comparison to Traditional Methods

The approach using Hill diversity and coverage standardization provides several advantages over traditional methods:

1.  **More robust to sampling differences**: By standardizing to equal coverage, we account for differences in the underlying species abundance distributions across sites and years.

2.  **Unified framework**: The Hill diversity framework provides a unified way to understand different aspects of diversity, from rare species (richness) to common species (Simpson).

3.  **Interpretable units**: All Hill diversity metrics are expressed in "effective number of species," making them more intuitive to interpret than traditional indices.

## References

Roswell, M., Dushoff, J., & Winfree, R. (2021). A conceptual guide to measuring species diversity. Oikos, 130(3), 321-338.
